---
title: "Lab 1: multiple linear regression"
subtitle: "Beer Consumption in Sao Paulo"
date: "Sept 6, 2019"
output: 
  tufte::tufte_html:
    tufte_variant: "envisioned"
    highlight: pygments
    css: lab.css
    toc: true
link-citations: yes
---

```{r include=FALSE}
library(tidyverse)
library(tufte)
library(knitr)
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  digits = 2
  )
knitr::opts_chunk$set(eval = FALSE)
```

**Due:** 6:00pm, Sept 6, 2019



# Introduction

The purpose of this lab is to give you additional practice working with multiple linear regression. You will also write your own code for doing k-fold cross validation. The lab is based on the Kaggle beer cosumption dataset found here:  https://www.kaggle.com/dongeorge/beer-consumption-sao-paulo/. Kaggle is a great online community of data scientists. To learn more about Kaggle, follow this link: https://www.kaggle.com/getting-started/44916.

In this lab you will demonstrate the impacts of variables on beer consumption in a given region and the consumption forecast for certain scenarios. The data (sample) were collected in São Paulo, Brazil, in a university area, where there are some parties with groups of students from 18 to 28 years of age (average).
   
<!-- # Packages -->

<!-- In this lab we will work with the `tidyverse` and `broom` packages. We can install and load them with the following: -->

<!-- ```{marginfigure} -->
<!-- When you install the tidyverse package, a long list of packages get installed with it. However when you load it (with the `library` function) only a few of them get loaded, e.g. `dplyr`, `ggplot2`, and `forcats`. The broom package is installed with the tidyverse, but we need to load it separately in order to make use of it. -->
<!-- ``` -->

<!-- ```{r eval = FALSE} -->
<!-- install.packages("tidyverse") -->
<!-- library(tidyverse)  -->
<!-- library(broom) -->
<!-- ``` -->



# Housekeeping

## R/RStudio

You all should have R and RStudio installed on your computers by now. If you do not, first install the latest version of R here: https://cran.rstudio.com (remember to select the right installer for your operating system). Next, install the latest version of RStudio here: https://www.rstudio.com/products/rstudio/download/. Scroll down to the "Installers for Supported Platforms" section and find the right installer for your operating system.

## R markdown

You are required to use R Markdown to type up this lab report. For a very (very!) basic R Markdown template, use this link: https://akandelanre.github.io/IDS702_F19/labs/resources/LabReport.Rmd

<!-- ## Create a GitHub account -->

<!-- Everyone should have a GitHub account by now. You won't need to use version control for this lab but you will need it for the team projects so you might as well create an account now. It is quite easy to use git/GitHub within RStudio If you have a GitHub account, skip the rest of this section.  -->

<!-- If you do not yet have a GitHub account, create one at https://github.com/. Try to incorporate your actual name and pick a name you will be comfortable revealing to your future boss! -->



<!-- # Warm up -->

<!-- **Pick one team member to complete the steps in this section while the others contribute to the discussion but do not actually touch the files on their computer.** -->

<!-- Before we introduce the data, let's warm up with some simple exercises. -->

<!-- ## YAML:  -->

<!-- Open the R Markdown (Rmd) file in your project, change the author name to your **team** name, and knit the document. -->

<!-- ## Commiting and pushing changes: -->

<!-- - Go to the **Git** pane in your RStudio.  -->
<!-- - View the **Diff** and confirm that you are happy with the changes. -->
<!-- - Add a commit message like "Update team name" in the **Commit message** box and hit **Commit**. -->
<!-- - Click on **Push**. This will prompt a dialogue box where you first need to enter your user name, and then your password. -->

<!-- ## Pulling changes: -->

<!-- Now, the remaining team members who have not been concurrently making these changes on their projects should click on the **Pull** button in their Git pane and observe that the changes are now reflected on their projects as well. -->


# The data

In this lab you will first download the data, then save it to the same directory as your R markdown file.

- Click [here](http://www2.stat.duke.edu/courses/Spring18/Sta199/labs/data/evals-mod.csv) to download the data. The file is called `evals-mod.csv`.
- Navigate to the data folder in your project and upload the `evals-mod.csv` file.

Load and clean the data by using the following R code.

```{marginfigure}
It is always a good idea to take a look at the first few rows of the raw file to see what the data looks like before loading the data. In this raw 'consumo_cerveja' file, you will notice that commas are actually used both as decimals and to separate the columns. Thus, you need to let R know by specifying the *sep* and *dec* options as in the code.
```

```{r eval = FALSE}
beer <- read.csv("consumo_cerveja.csv",stringsAsFactors = FALSE, sep = ",",dec=",")
# rename the variables
beer$date <- beer$Data
beer$temp_median_c <- beer$Temperatura.Media..C.
beer$temp_min_c <- beer$Temperatura.Minima..C.
beer$temp_max_c <- beer$Temperatura.Maxima..C.
beer$precip_mm <- beer$Precipitacao..mm.
beer$weekend <- factor(beer$Final.de.Semana)
beer$beer_cons_liters <- as.numeric(beer$Consumo.de.cerveja..litros.)
beer <- beer[ , 8:ncol(beer)]
```

After renaming the variables using the code above, 

# Exercises

Treat the variable `beer_cons_liters` as your response variable and the other variables as potential predictors.

1. Make a histogram of `beer_cons_liters`. Describe the distribution. Does it look normal or skewed? Will the normality assumption be a plausible one here?

2. Make exploratory plots of `beer_cons_liters` versus each potential predictor. Are all the relationships linear? If any one of them is nonlinear, describe the distribution.

3. Fit a linear model using `weekend`, `precip_mm`, and `temp_median_c` as your predictors. Interpret your results. Which of the variables would is the best covariate for predicting beer consumption?

4. Compute the in-sample RMSE for model in question 3.

5. Write a code for doing k-fold cross validation. Let $k=10$ and use average RMSE as the metric for quantifying predictive error. What is the average RMSE for the model in question 3 above?

6. Fit another model that includes interactions between `weekend` and the other two predictors. Are the interaction terms significant?

7. Use your code for the k-fold cross validation to compute the average RMSE for this new model in question 6. Is the new RMSE model lower or higher? What can you infer from that?




## Part 1: Simple linear regression

2. Fit a linear model (one you have fit before): `m_bty`, predicting average
   professor evaluation `score` based on average beauty rating (`bty_avg`) only. Write the 
   linear model, and note the $R^2$ and the adjusted $R^2$.

## Part 2: Multiple linear regression

2. Fit a linear model (one you have fit before): `m_bty_gen`, predicting average
   professor evaluation `score` based on average beauty rating (`bty_avg`) and `gender`. 
   Write the linear model, and note the $R^2$ and the adjusted $R^2$.
   
3. Interpret the slope and intercept of `m_bty_gen` in context of the data.

4. What percent of the variability in `score` is explained by the model `m_bty_gen`.

5.  What is the equation of the line corresponding to *just* male professors?
    
6.  For two professors who received the same beauty rating, which gender tends 
    to have the higher course evaluation score?
    
7. How does the relationship between beauty and evaluation score
    vary between male and female professors?
    
8. How do the adjusted $R^2$ values of `m_bty_gen` and `m_bty` compare? What does this tell us 
   about how useful `gender` is in explaining the variability in evaluation scores when we 
   already have information on the beaty score of the professor.

9. Compare the slopes of `bty_avg` under the two models (`m_bty` and `m_bty_gen`). Has the 
   addition of `gender` to the model changed the parameter estimate (slope) for `bty_avg`?
    
10. Create a new model called `m_bty_rank` with `gender` removed and `rank` 
    added in. Write the equation of the linear model and interpret the slopes and intercept 
    in context of the data. 

## Part 5: The search for the best model

Going forward, only consider the following variables as potential predictors: `rank`, 
`ethnicity`, `gender`, `language`, `age`, `cls_perc_eval`, `cls_did_eval`, `cls_students`, 
`cls_level`, `cls_profs`, `cls_credits`, `bty_avg`.

11. Which variable, on its own, would you expect to be the worst predictor of 
    evaluation scores? Why? *Hint:* Think about which variable would you 
    expect to not have any association with the professor's score.

12. Check your suspicions from the previous exercise. Include the model output
    for that variable in your response.
    
13. Suppose you wanted to fit a full model with the variables listed above. If 
    you are already going to include `cls_perc_eval` and `cls_students`, which variable 
    should you not include as an additional predictor? Why?

14. Fit a full model with all predictors listed above (except for the one you decided to 
    exclude) in the previous question.

15. Using backward-selection with adjusted R-squared as the selection 
    criterion, determine the *best* model. You do not need to show all 
    steps in your answer, just the output for the final model. Also, 
    write out the linear model for predicting score based on the final 
    model you settle on.

16. Interpret the slopes of one numerical and one categorical predictor based on your final model.

16. Based on your final model, describe the characteristics of a professor and 
    course at University of Texas at Austin that would be associated with a high
    evaluation score.

17. Would you be comfortable generalizing your conclusions to apply to professors
    generally (at any university)? Why or why not?
    
    
### Acknowledgement

This lab includes ideas proposed by [Sam Voisin](https://stat.duke.edu/people/sam-voisin).
    